\begin{acknowledgments}
	I would like to thank several people who supported me throughout the process of completing my thesis.

	I'm grateful to Dr.~Schooley, who was patient, supportive, and gave great direction throughout the process while letting me explore and dig into what interested me.
	Thank you for trusting me to see this through.
	I'm grateful for Dr.~Hansen, Dr.~Tay, and Audra McHoes for their help and faith in me.
	Thank you to Mary Hedengren for listening and helping me improve my writing.
	Thank you to Dr.~Page for welcoming me into your lab and teaching me that academic research can be exciting and impactful.
	Thank you to Dr.~Seamons for the great discussions on usable security which inspired me and gave direction for this research.

	I'd like to acknowledge the CyberCorps Scholarship for Service program for funding and enabling this research, and the support from BYU Graduate Studies and the BYU Electrical and Computer Engineering Department that made it possible for me to publish and present my work.

	Thank you to my BYU peers, friends, and coworkers at Paramify who checked in on me, volunteered to participate in the user study, and encouraged me to keep going.
	I'm especially grateful to Kenny Scott, Mike Cook, and the whole team at Paramify for letting me work flexible hours to finish my degree.
	Here's to never having another password-protected SSP!

	Most of all, thank you to my family.
	To my parents, for encouraging me to ``get as much education as you can'' and for always believing I can do great things.
	And to my wife Anne, who loved, encouraged, and supported me through all the ups and downs.
	I love you!

	Lastly, I'm thankful for the singular opportunity to study at BYU and seek learning by study and by faith.
	I will be forever grateful for what I've learned and who I've become.
\end{acknowledgments}
